Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-01-27 manuscript.tex
latest

Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$/$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.

The importance of $b$ in securing territories is a general feature of lottery competition. Indeed, as can be seen from Eq. \eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}. This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists. Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than coexistence in a single, stable environment \citep{chesson_1981}. It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.

Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules , regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

>>>>

Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, the $r$/$K$ dichotomy is also not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.

The importance of $b$ at high densities is a consequence of the unpredictable nature of competition at the territorial level. Even in the extreme case that one genotype is guaranteed to defeat all others in a ``fair'' contest (e.g. it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win any given territory by sheer luck. This could be because of which propagules happen to arrive first (a dispersal advantage), or other stochastic factors. The lottery model of competition captures this effect. Viewed through the lens of the Wright-Fisher model, the importance of $b$ at high densities is obvious --- the analogous effect is that fecundity confers a relative fitness advantage even though the population is not growing.

In the classic lottery model, $b_i$ and $c_i$ are essentially equivalent in that only the product $b_i c_i$ matters (see ``Model''). This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where there is a colonization-competition trade-off \citep{tilman_94}, and co-existence between $b$ and $c$ specialists.

It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.
